---
title: "AB01 Interval Maturation Scoring"
description: "PCA views and D21-anchored interval maturation scores for AB01."
output:
  html_document:
    toc: true
    toc_depth: 2
    code_folding: none
    df_print: paged
---

```{r setup}
# Created:
# 2026-06-03
#
# Inputs:
# - data/AB01/metadata/WC24_metadata_clean.rds: AB01 sample metadata
# - data/AB01/raw counts/WC24_filt_raw_counts.rds: AB01 raw count matrix
# - data_DSHC1/ZRN01/metadata/ZRN01_metadata.rds: ZRN01 sample metadata
# - data_DSHC1/ZRN01/counts/ZRN01_vst.rds: ZRN01 VST matrix
#
# Outputs:
# - docs/AB01_interval_maturation_score.html: rendered standalone HTML report
#
# Purpose:
# Test D21-anchored PCA maturation scoring in AB01 and show complementary PCA
# views using D21 samples at D6, D10, and D17.
#
# Notes:
# VST is computed exactly once with DESeq2::vst(blind = FALSE) using the
# genotype_timepoint condition after CPM-based expression filtering. The source
# metadata uses D21 and T21 labels; figures display those groups as Di21 and
# Tri21 for readability.

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = TRUE,
  fig.align = 'center',
  dpi = 600)

project_root <- if (base::basename(base::getwd()) == 'scripts') {
  base::normalizePath('..')
} else {
  base::getwd()
}
knitr::opts_knit$set(root.dir = project_root)

required_packages <- base::c(
  'DESeq2',
  'SummarizedExperiment',
  'ggplot2',
  'grid')
missing_packages <- required_packages[
  !base::vapply(
    required_packages,
    base::requireNamespace,
    base::logical(1),
    quietly = TRUE)
]
if (base::length(missing_packages) > 0L) {
  base::stop(
    'missing required packages: ',
    base::paste(missing_packages, collapse = ', '))
}

plot_font_family <- 'Arial'

theme_ab01 <- function(base_size = 10) {
  ggplot2::theme_classic(
    base_size = base_size,
    base_family = plot_font_family) +
    ggplot2::theme(
      text = ggplot2::element_text(color = 'black', family = plot_font_family),
      axis.text = ggplot2::element_text(color = 'black', family = plot_font_family),
      axis.title = ggplot2::element_text(color = 'black', family = plot_font_family),
      axis.line = ggplot2::element_line(color = 'black', linewidth = 0.3),
      axis.ticks = ggplot2::element_line(color = 'black', linewidth = 0.3),
      panel.grid.major.y = ggplot2::element_line(color = 'grey92', linewidth = 0.2),
      panel.grid.major.x = ggplot2::element_blank(),
      panel.grid.minor = ggplot2::element_blank(),
      strip.background = ggplot2::element_blank(),
      strip.text = ggplot2::element_text(
        color = 'black',
        face = 'bold',
        family = plot_font_family),
      legend.title = ggplot2::element_blank(),
      legend.text = ggplot2::element_text(color = 'black', family = plot_font_family))
}
```

# Overview

This report evaluates AB01 maturation using D21 samples as the reference
trajectory. It adds two D21-only PCA views before scoring:

- one PCA trained on the top 1000 most variable genes across D21 D6, D10, and D17
  samples together;
- one PCA trained on the union of the top 1000 D21-variable genes from the D6-D10
  and D10-D17 adjacent intervals.

The score plots use the first three PCs from each interval-specific PCA. For
D6-D10, the D21 D6 centroid is anchored at 0 and the D21 D10 centroid is anchored
at 1. For D10-D17, the D21 D10 centroid is anchored at 1 and the D21 D17 centroid
is anchored at 2. T21 samples are projected into the same fixed D21-trained PCA
space and scored relative to those D21 anchors.

# Load AB01 Data

The metadata and count matrix are loaded directly from the declared AB01 inputs.
The stored count columns are reordered to metadata order before any expression
filtering or VST calculation.

```{r load-data}
metadata <- base::readRDS('data/AB01/metadata/WC24_metadata_clean.rds')
counts <- base::readRDS('data/AB01/raw counts/WC24_filt_raw_counts.rds')

base::stopifnot(!base::anyDuplicated(metadata$sample_id))
base::stopifnot(base::all(metadata$sample_id %in% base::colnames(counts)))

counts <- counts[, metadata$sample_id, drop = FALSE]
base::stopifnot(base::identical(
  base::as.character(metadata$sample_id),
  base::colnames(counts)))

metadata$timepoint_numeric <- base::as.numeric(base::as.character(metadata$timepoint))
timepoint_levels <- base::paste0(
  'D',
  base::sort(base::unique(metadata$timepoint_numeric)))
metadata$timepoint_label <- base::factor(
  base::paste0('D', metadata$timepoint_numeric),
  levels = timepoint_levels)
metadata$genotype <- base::factor(metadata$genotype, levels = base::c('D21', 'T21'))
metadata$genotype_display <- base::factor(
  base::as.character(metadata$genotype),
  levels = base::c('D21', 'T21'),
  labels = base::c('Di21', 'Tri21'))
metadata$genotype_timepoint <- base::factor(
  metadata$genotype_timepoint,
  levels = base::unique(metadata$genotype_timepoint[
    base::order(metadata$timepoint_numeric, metadata$genotype)
  ]))
base::rownames(metadata) <- metadata$sample_id

input_summary <- base::data.frame(
  samples = base::nrow(metadata),
  genotypes = base::paste(base::levels(metadata$genotype_display), collapse = ', '),
  timepoints = base::paste(
    base::sort(base::unique(metadata$timepoint_numeric)),
    collapse = ', '),
  stringsAsFactors = FALSE)

input_summary
```

# Preprocess Counts and Compute VST Once

The CPM filter removes genes with minimal evidence of expression before VST.
This is intentionally done once so every PCA and score uses the same transformed
expression matrix.

```{r compute-vst}
# Critical expression filter: retain genes with CPM >= 1 in at least 3 samples.
library_sizes <- base::colSums(counts)
cpm_matrix <- base::sweep(counts, 2, library_sizes, FUN = '/') * 1e6
keep_genes <- base::rowSums(cpm_matrix >= 1) >= 3
filtered_counts <- counts[keep_genes, , drop = FALSE]

preprocessing_summary <- base::data.frame(
  retained_genes_after_cpm_filter = base::nrow(filtered_counts),
  cpm_cutoff = 1,
  minimum_samples = 3,
  stringsAsFactors = FALSE)

preprocessing_summary

dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = filtered_counts,
  colData = metadata,
  design = ~genotype_timepoint)

vst_matrix <- SummarizedExperiment::assay(
  DESeq2::vst(dds, blind = FALSE))

base::stopifnot(base::identical(base::colnames(vst_matrix), metadata$sample_id))
```

# Define PCA and Scoring Helpers

These helpers keep gene selection, PCA projection, variance labeling, and
interval scoring explicit. PCA is centered but not scaled, so genes with larger
VST variation contribute more strongly after the D21-only feature-selection step.

```{r define-functions}
# Critical PCA parameter: all requested PCA gene sets use the top 1000 genes.
pca_gene_count = 1000L

# Critical scoring parameter: maturation scores use PC1, PC2, and PC3.
n_scoring_pcs = 3L

# Sensitivity parameter: repeat scoring every 100 genes through the full filtered VST gene set.
sensitivity_gene_counts <- base::unique(base::c(
  base::seq(100L, base::nrow(vst_matrix), by = 100L),
  base::nrow(vst_matrix)))
sensitivity_x_breaks <- base::seq(500L, base::max(sensitivity_gene_counts), by = 500L)

interval_specs <- base::data.frame(
  interval = base::c('D6-D10', 'D10-D17'),
  start_timepoint = base::c(6, 10),
  end_timepoint = base::c(10, 17),
  score_start = base::c(0, 1),
  score_end = base::c(1, 2),
  stringsAsFactors = FALSE)
interval_specs$interval <- base::factor(
  interval_specs$interval,
  levels = base::c('D6-D10', 'D10-D17'))

genotype_colors <- base::c(Di21 = '#2563EB', Tri21 = '#DC2626')
timepoint_colors <- base::c(
  D4 = '#64748B',
  D6 = '#059669',
  D7 = '#0D9488',
  D10 = '#7C3AED',
  D15 = '#F59E0B',
  D17 = '#EA580C')

get_sample_ids <- function(sample_metadata, timepoints, genotype = NULL) {
  keep <- sample_metadata$timepoint_numeric %in% timepoints
  if (!base::is.null(genotype)) {
    keep <- keep & sample_metadata$genotype == genotype
  }
  base::as.character(sample_metadata$sample_id[keep])
}

select_top_variable_genes <- function(vst_matrix, sample_metadata, timepoints,
                                      genotype = 'D21',
                                      requested_gene_count = pca_gene_count) {
  sample_ids <- get_sample_ids(
    sample_metadata = sample_metadata,
    timepoints = timepoints,
    genotype = genotype)
  gene_variance <- base::apply(vst_matrix[, sample_ids, drop = FALSE], 1, stats::var)
  ranked_genes <- base::names(base::sort(gene_variance, decreasing = TRUE))
  effective_gene_count <- base::min(requested_gene_count, base::length(ranked_genes))

  base::data.frame(
    gene_id = ranked_genes[base::seq_len(effective_gene_count)],
    variance = gene_variance[ranked_genes[base::seq_len(effective_gene_count)]],
    requested_gene_count = requested_gene_count,
    effective_gene_count = effective_gene_count,
    stringsAsFactors = FALSE)
}

project_samples_into_pca <- function(vst_subset, pca_fit) {
  if (!base::identical(base::rownames(vst_subset), base::rownames(pca_fit$rotation))) {
    base::stop('vst_subset rows must match pca_fit rotation rows exactly.')
  }

  projected <- base::scale(
    base::t(vst_subset),
    center = pca_fit$center,
    scale = FALSE)
  coord <- base::as.data.frame(projected %*% pca_fit$rotation)
  coord$sample_id <- base::rownames(coord)
  coord
}

calculate_variance_explained <- function(pca_fit) {
  variance_explained <- (pca_fit$sdev^2) / base::sum(pca_fit$sdev^2) * 100
  stats::setNames(variance_explained, base::paste0('PC', base::seq_along(variance_explained)))
}

format_pc_axis_label <- function(pc_name, variance_explained) {
  base::paste0(pc_name, ' (', base::round(variance_explained[[pc_name]], 1), '% variance)')
}

build_d21_pca <- function(vst_matrix, sample_metadata, gene_ids, training_timepoints) {
  training_ids <- get_sample_ids(
    sample_metadata = sample_metadata,
    timepoints = training_timepoints,
    genotype = 'D21')
  pca_fit <- stats::prcomp(
    base::t(vst_matrix[gene_ids, training_ids, drop = FALSE]),
    center = TRUE,
    scale. = FALSE)
  coords <- project_samples_into_pca(
    vst_subset = vst_matrix[gene_ids, training_ids, drop = FALSE],
    pca_fit = pca_fit)
  coords <- base::merge(
    sample_metadata,
    coords,
    by = 'sample_id',
    sort = FALSE)
  coords <- coords[base::match(training_ids, coords$sample_id), , drop = FALSE]

  base::list(
    pca_fit = pca_fit,
    coords = coords,
    variance_explained = calculate_variance_explained(pca_fit))
}

plot_d21_pca <- function(pca_result, plot_title) {
  ggplot2::ggplot(
    pca_result$coords,
    ggplot2::aes(PC1, PC2, color = timepoint_label)) +
    ggplot2::geom_point(size = 2.8, stroke = 0.8) +
    ggplot2::scale_color_manual(values = timepoint_colors) +
    ggplot2::labs(
      x = format_pc_axis_label('PC1', pca_result$variance_explained),
      y = format_pc_axis_label('PC2', pca_result$variance_explained),
      title = plot_title) +
    theme_ab01(base_size = 10)
}

plot_scree <- function(pca_result, plot_title, n_pcs = 15L) {
  n_available_pcs <- base::min(n_pcs, base::length(pca_result$variance_explained))
  scree_df <- base::data.frame(
    pc = base::factor(
      base::paste0('PC', base::seq_len(n_available_pcs)),
      levels = base::paste0('PC', base::seq_len(n_available_pcs))),
    variance_explained = pca_result$variance_explained[base::seq_len(n_available_pcs)],
    stringsAsFactors = FALSE)

  ggplot2::ggplot(
    scree_df,
    ggplot2::aes(pc, variance_explained)) +
    ggplot2::geom_col(width = 0.72, fill = '#4B5563', color = 'black', linewidth = 0.25) +
    ggplot2::labs(
      x = NULL,
      y = '% variance explained',
      title = plot_title) +
    theme_ab01(base_size = 10) +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
}

score_interval <- function(vst_matrix, sample_metadata, interval_row,
                           requested_gene_count = pca_gene_count,
                           n_score_pcs = n_scoring_pcs) {
  start_timepoint <- interval_row$start_timepoint[[1]]
  end_timepoint <- interval_row$end_timepoint[[1]]
  score_start <- interval_row$score_start[[1]]
  score_end <- interval_row$score_end[[1]]

  gene_table <- select_top_variable_genes(
    vst_matrix = vst_matrix,
    sample_metadata = sample_metadata,
    timepoints = base::c(start_timepoint, end_timepoint),
    genotype = 'D21',
    requested_gene_count = requested_gene_count)
  gene_ids <- gene_table$gene_id

  training_ids <- get_sample_ids(
    sample_metadata = sample_metadata,
    timepoints = base::c(start_timepoint, end_timepoint),
    genotype = 'D21')
  projection_ids <- get_sample_ids(
    sample_metadata = sample_metadata,
    timepoints = base::c(start_timepoint, end_timepoint))

  pca_fit <- stats::prcomp(
    base::t(vst_matrix[gene_ids, training_ids, drop = FALSE]),
    center = TRUE,
    scale. = FALSE)
  pc_cols <- base::paste0(
    'PC',
    base::seq_len(base::min(n_score_pcs, base::ncol(pca_fit$x))))

  coords <- project_samples_into_pca(
    vst_subset = vst_matrix[gene_ids, projection_ids, drop = FALSE],
    pca_fit = pca_fit)
  coords <- coords[, base::c('sample_id', pc_cols), drop = FALSE]
  score_df <- base::merge(
    sample_metadata,
    coords,
    by = 'sample_id',
    sort = FALSE)
  score_df <- score_df[base::match(projection_ids, score_df$sample_id), , drop = FALSE]

  start_centroid <- base::colMeans(score_df[
    score_df$genotype == 'D21' & score_df$timepoint_numeric == start_timepoint,
    pc_cols,
    drop = FALSE
  ])
  end_centroid <- base::colMeans(score_df[
    score_df$genotype == 'D21' & score_df$timepoint_numeric == end_timepoint,
    pc_cols,
    drop = FALSE
  ])
  vector_raw <- end_centroid - start_centroid
  vector_length_sq <- base::sum(vector_raw^2)
  if (!base::is.finite(vector_length_sq) || vector_length_sq <= 0) {
    base::stop('interval centroid vector has zero length.')
  }

  coords_mat <- base::as.matrix(score_df[, pc_cols, drop = FALSE])
  centered_coords <- coords_mat - base::matrix(
    start_centroid,
    nrow = base::nrow(coords_mat),
    ncol = base::length(pc_cols),
    byrow = TRUE)
  unit_interval_score <- base::as.numeric(centered_coords %*% vector_raw) / vector_length_sq
  score_df$maturation_score <- score_start + unit_interval_score * (score_end - score_start)

  score_df$interval <- interval_row$interval[[1]]
  score_df$interval <- base::factor(
    score_df$interval,
    levels = base::levels(interval_specs$interval))
  score_df$interval_start_timepoint <- start_timepoint
  score_df$interval_end_timepoint <- end_timepoint
  score_df$score_start <- score_start
  score_df$score_end <- score_end
  score_df$requested_gene_count <- requested_gene_count
  score_df$effective_gene_count <- gene_table$effective_gene_count[[1]]

  base::list(
    score_table = score_df,
    gene_table = gene_table,
    pca_fit = pca_fit,
    pc_columns = pc_cols,
    start_centroid = start_centroid,
    end_centroid = end_centroid,
    variance_explained = calculate_variance_explained(pca_fit))
}
```

# D21 PCA Across D6, D10, and D17

This PCA asks whether D21 samples separate by developmental day when the feature
set is selected from all three D21 timepoints together. Genes are ranked by VST
variance across D21 D6, D10, and D17 samples, and the top 1000 genes are used for
PCA.

```{r all-timepoint-d21-pca}
all_timepoint_gene_table <- select_top_variable_genes(
  vst_matrix = vst_matrix,
  sample_metadata = metadata,
  timepoints = base::c(6, 10, 17),
  genotype = 'D21',
  requested_gene_count = pca_gene_count)

all_timepoint_pca <- build_d21_pca(
  vst_matrix = vst_matrix,
  sample_metadata = metadata,
  gene_ids = all_timepoint_gene_table$gene_id,
  training_timepoints = base::c(6, 10, 17))

base::data.frame(
  requested_gene_count = all_timepoint_gene_table$requested_gene_count[[1]],
  effective_gene_count = all_timepoint_gene_table$effective_gene_count[[1]],
  largest_gene_variance = base::max(all_timepoint_gene_table$variance),
  stringsAsFactors = FALSE)
```

```{r all-timepoint-d21-pca-plot, fig.width=5.2, fig.height=4.4}
plot_d21_pca(
  pca_result = all_timepoint_pca,
  plot_title = 'D21 PCA using top variable genes across D6, D10, and D17')
```

```{r all-timepoint-d21-scree-plot, fig.width=6.2, fig.height=4.2}
plot_scree(
  pca_result = all_timepoint_pca,
  plot_title = 'D21 all-timepoint PCA scree plot')
```

# D21 PCA Using Union of Adjacent-Interval Genes

This PCA uses a union feature set from two adjacent D21 comparisons. The first
set is the top 1000 most variable genes across D21 D6 and D10 samples. The
second set is the top 1000 most variable genes across D21 D10 and D17 samples.
The union is then used to train one D21 PCA across D6, D10, and D17.

```{r union-interval-d21-pca}
d6_d10_gene_table <- select_top_variable_genes(
  vst_matrix = vst_matrix,
  sample_metadata = metadata,
  timepoints = base::c(6, 10),
  genotype = 'D21',
  requested_gene_count = pca_gene_count)
d10_d17_gene_table <- select_top_variable_genes(
  vst_matrix = vst_matrix,
  sample_metadata = metadata,
  timepoints = base::c(10, 17),
  genotype = 'D21',
  requested_gene_count = pca_gene_count)

union_interval_genes <- base::union(
  d6_d10_gene_table$gene_id,
  d10_d17_gene_table$gene_id)

union_interval_pca <- build_d21_pca(
  vst_matrix = vst_matrix,
  sample_metadata = metadata,
  gene_ids = union_interval_genes,
  training_timepoints = base::c(6, 10, 17))

base::data.frame(
  d6_d10_genes = base::length(d6_d10_gene_table$gene_id),
  d10_d17_genes = base::length(d10_d17_gene_table$gene_id),
  union_genes = base::length(union_interval_genes),
  overlap_genes = base::length(base::intersect(d6_d10_gene_table$gene_id, d10_d17_gene_table$gene_id)),
  stringsAsFactors = FALSE)
```

```{r union-interval-d21-pca-plot, fig.width=5.2, fig.height=4.4}
plot_d21_pca(
  pca_result = union_interval_pca,
  plot_title = 'D21 PCA using union of adjacent-interval variable genes')
```

```{r union-interval-d21-scree-plot, fig.width=6.2, fig.height=4.2}
plot_scree(
  pca_result = union_interval_pca,
  plot_title = 'D21 union-gene PCA scree plot')
```

# D21-Anchored Interval Scores

Each interval score is computed in its own D21-trained PCA space. The feature
set is selected from the D21 samples in that interval only, the PCA is trained on
those D21 samples, and both D21 and T21 samples from the same two timepoints are
projected into that fixed PCA. The maturation coordinate is the projection of
each sample onto the D21 start-to-end centroid vector using PC1, PC2, and PC3.

```{r run-primary-score}
primary_results <- base::lapply(
  base::seq_len(base::nrow(interval_specs)),
  function(i) {
    score_interval(
      vst_matrix = vst_matrix,
      sample_metadata = metadata,
      interval_row = interval_specs[i, , drop = FALSE],
      requested_gene_count = pca_gene_count,
      n_score_pcs = n_scoring_pcs)
  })
base::names(primary_results) <- interval_specs$interval

primary_scores <- base::do.call(
  base::rbind,
  base::lapply(primary_results, function(x) x$score_table))
base::rownames(primary_scores) <- NULL

primary_scores[
  ,
  base::c(
    'interval',
    'sample_id',
    'genotype_display',
    'timepoint_label',
    'rep',
    'PC1',
    'PC2',
    'PC3',
    'maturation_score'),
  drop = FALSE
]
```

## Interval PCA Views

These PCA plots show the fixed PCA spaces used for scoring. Axis labels include
the variance explained by PC1 and PC2 for each interval-specific PCA.

```{r define-interval-pca-plot}
plot_interval_pca <- function(interval_name) {
  interval_scores <- primary_scores[primary_scores$interval == interval_name, , drop = FALSE]
  variance_explained <- primary_results[[interval_name]]$variance_explained

  ggplot2::ggplot(
    interval_scores,
    ggplot2::aes(PC1, PC2, color = timepoint_label, shape = genotype_display)) +
    ggplot2::geom_point(size = 2.6, stroke = 0.8) +
    ggplot2::scale_color_manual(values = timepoint_colors) +
    ggplot2::scale_shape_manual(values = base::c(Di21 = 16, Tri21 = 17)) +
    ggplot2::labs(
      x = format_pc_axis_label('PC1', variance_explained),
      y = format_pc_axis_label('PC2', variance_explained),
      title = base::paste0(interval_name, ' D21-trained PCA space')) +
    theme_ab01(base_size = 10)
}

interval_pca_scores <- primary_scores
interval_pca_scores$interval <- base::factor(
  interval_pca_scores$interval,
  levels = base::levels(interval_specs$interval))
```

```{r d6-d10-interval-pca-plot, fig.width=5.4, fig.height=4.4}
plot_interval_pca('D6-D10')
```

```{r d10-d17-interval-pca-plot, fig.width=5.4, fig.height=4.4}
plot_interval_pca('D10-D17')
```

## Individual-Day Score Plots

The bars show mean maturation score for each genotype at each interval timepoint,
with individual replicates overlaid. Within a timepoint, Di21 and Tri21 bars are
paired directly beside each other; gaps separate the two days in the interval.

```{r score-plot-data}
score_mean <- stats::aggregate(
  maturation_score ~ interval + genotype_display + timepoint_label + timepoint_numeric,
  data = primary_scores,
  FUN = base::mean)

score_points <- primary_scores

score_mean$timepoint_index <- NA_integer_
score_points$timepoint_index <- NA_integer_
for (interval_name in base::levels(interval_specs$interval)) {
  interval_timepoints <- base::sort(base::unique(score_mean$timepoint_numeric[score_mean$interval == interval_name]))
  score_mean$timepoint_index[score_mean$interval == interval_name] <- base::match(
    score_mean$timepoint_numeric[score_mean$interval == interval_name],
    interval_timepoints)
  score_points$timepoint_index[score_points$interval == interval_name] <- base::match(
    score_points$timepoint_numeric[score_points$interval == interval_name],
    interval_timepoints)
}

score_mean$plot_x <- base::ifelse(
  score_mean$genotype_display == 'Di21',
  score_mean$timepoint_index - 0.18,
  score_mean$timepoint_index + 0.18)
score_points$plot_x <- base::ifelse(
  score_points$genotype_display == 'Di21',
  score_points$timepoint_index - 0.18,
  score_points$timepoint_index + 0.18)

score_mean[
  base::order(score_mean$interval, score_mean$timepoint_numeric, score_mean$genotype_display),
  ,
  drop = FALSE
]
```

```{r d6-d10-score-plot, fig.width=5.8, fig.height=4.2}
d6_d10_score_mean <- score_mean[score_mean$interval == 'D6-D10', , drop = FALSE]
d6_d10_score_points <- score_points[score_points$interval == 'D6-D10', , drop = FALSE]

ggplot2::ggplot(
  d6_d10_score_mean,
  ggplot2::aes(plot_x, maturation_score, fill = genotype_display)) +
  ggplot2::geom_hline(
    yintercept = base::c(0, 1),
    linetype = 'dashed',
    color = 'grey45',
    linewidth = 0.25) +
  ggplot2::geom_col(
    width = 0.36,
    color = 'black',
    linewidth = 0.25,
    alpha = 0.9) +
  ggplot2::geom_point(
    data = d6_d10_score_points,
    ggplot2::aes(plot_x, maturation_score, fill = genotype_display),
    inherit.aes = FALSE,
    size = 1.9,
    shape = 21,
    color = 'black',
    stroke = 0.25,
    position = ggplot2::position_jitter(width = 0.04, height = 0, seed = 42)) +
  ggplot2::scale_x_continuous(
    breaks = base::sort(base::unique(d6_d10_score_mean$timepoint_index)),
    labels = base::paste0('D', base::sort(base::unique(d6_d10_score_mean$timepoint_numeric))),
    expand = ggplot2::expansion(mult = c(0.12, 0.12))) +
  ggplot2::scale_fill_manual(values = genotype_colors) +
  ggplot2::labs(
    x = NULL,
    y = 'Maturation score',
    title = 'D6-D10 score anchored to Di21 centroids') +
  theme_ab01(base_size = 10)
```

```{r d10-d17-score-plot, fig.width=5.8, fig.height=4.2}
d10_d17_score_mean <- score_mean[score_mean$interval == 'D10-D17', , drop = FALSE]
d10_d17_score_points <- score_points[score_points$interval == 'D10-D17', , drop = FALSE]

ggplot2::ggplot(
  d10_d17_score_mean,
  ggplot2::aes(plot_x, maturation_score, fill = genotype_display)) +
  ggplot2::geom_hline(
    yintercept = base::c(1, 2),
    linetype = 'dashed',
    color = 'grey45',
    linewidth = 0.25) +
  ggplot2::geom_col(
    width = 0.36,
    color = 'black',
    linewidth = 0.25,
    alpha = 0.9) +
  ggplot2::geom_point(
    data = d10_d17_score_points,
    ggplot2::aes(plot_x, maturation_score, fill = genotype_display),
    inherit.aes = FALSE,
    size = 1.9,
    shape = 21,
    color = 'black',
    stroke = 0.25,
    position = ggplot2::position_jitter(width = 0.04, height = 0, seed = 42)) +
  ggplot2::scale_x_continuous(
    breaks = base::sort(base::unique(d10_d17_score_mean$timepoint_index)),
    labels = base::paste0('D', base::sort(base::unique(d10_d17_score_mean$timepoint_numeric))),
    expand = ggplot2::expansion(mult = c(0.12, 0.12))) +
  ggplot2::scale_fill_manual(values = genotype_colors) +
  ggplot2::labs(
    x = NULL,
    y = 'Maturation score',
    title = 'D10-D17 score anchored to Di21 centroids') +
  theme_ab01(base_size = 10)
```

# Sensitivity of Scores to Gene-List Size

The sensitivity analysis repeats the same D21-anchored scoring workflow across a
range of input gene-list sizes. Each run reselects the most variable D21 genes
for the relevant interval, retrains the interval PCA, and recomputes scores from
PC1, PC2, and PC3. This checks whether the genotype and timepoint score patterns
depend strongly on choosing exactly 1000 genes.

```{r run-sensitivity}
sensitivity_results <- base::list()
result_idx <- 1L

for (interval_idx in base::seq_len(base::nrow(interval_specs))) {
  for (gene_count in sensitivity_gene_counts) {
    sensitivity_results[[result_idx]] <- score_interval(
      vst_matrix = vst_matrix,
      sample_metadata = metadata,
      interval_row = interval_specs[interval_idx, , drop = FALSE],
      requested_gene_count = gene_count,
      n_score_pcs = n_scoring_pcs)
    result_idx <- result_idx + 1L
  }
}

sensitivity_scores <- base::do.call(
  base::rbind,
  base::lapply(sensitivity_results, function(x) x$score_table))
base::rownames(sensitivity_scores) <- NULL

sensitivity_summary <- stats::aggregate(
  maturation_score ~ interval + requested_gene_count + effective_gene_count +
    genotype_display + timepoint_numeric + timepoint_label,
  data = sensitivity_scores,
  FUN = base::mean)

sensitivity_summary <- sensitivity_summary[
  base::order(
    sensitivity_summary$interval,
    sensitivity_summary$requested_gene_count,
    sensitivity_summary$timepoint_numeric,
    sensitivity_summary$genotype_display),
  ,
  drop = FALSE]

utils::head(sensitivity_summary, 12)
```

## Mean Score Across Gene Counts

This plot shows the mean score for each genotype and timepoint as the number of
selected D21-variable genes increases. Stable lines indicate that the score is
not highly sensitive to one exact gene-count cutoff.

```{r sensitivity-mean-score, fig.width=9.5, fig.height=5.2}
sensitivity_summary$label_group <- base::factor(
  base::paste(sensitivity_summary$genotype_display, sensitivity_summary$timepoint_label),
  levels = base::c(
    'Di21 D6',
    'Tri21 D6',
    'Di21 D10',
    'Tri21 D10',
    'Di21 D17',
    'Tri21 D17'))

ggplot2::ggplot(
  sensitivity_summary,
  ggplot2::aes(
    effective_gene_count,
    maturation_score,
    color = genotype_display,
    linetype = timepoint_label,
    group = label_group)) +
  ggplot2::geom_hline(
    yintercept = base::c(0, 1, 2),
    linetype = 'dashed',
    color = 'grey45',
    linewidth = 0.25) +
  ggplot2::geom_line(linewidth = 0.62) +
  ggplot2::geom_point(size = 0.9) +
  ggplot2::facet_wrap(~interval, scales = 'free_y') +
  ggplot2::scale_x_continuous(breaks = sensitivity_x_breaks) +
  ggplot2::scale_color_manual(values = genotype_colors) +
  ggplot2::labs(
    x = 'Effective number of input genes',
    y = 'Mean maturation score',
    title = 'Sensitivity of genotype-timepoint mean scores to gene-list size') +
  theme_ab01(base_size = 10) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

## Tri21 Displacement From Matched Di21

This plot subtracts the matched Di21 mean score from the Tri21 mean score at the
same interval, gene count, and timepoint. It is a compact way to inspect whether
Tri21 samples score earlier or later than their Di21 timepoint anchor.

```{r sensitivity-tri21-displacement, fig.width=9, fig.height=4.8}
d21_reference <- sensitivity_summary[
  sensitivity_summary$genotype_display == 'Di21',
  base::c(
    'interval',
    'requested_gene_count',
    'effective_gene_count',
    'timepoint_numeric',
    'maturation_score'),
  drop = FALSE]
base::names(d21_reference)[base::names(d21_reference) == 'maturation_score'] <- 'd21_mean_score'

tri21_displacement <- base::merge(
  sensitivity_summary[sensitivity_summary$genotype_display == 'Tri21', , drop = FALSE],
  d21_reference,
  by = base::c(
    'interval',
    'requested_gene_count',
    'effective_gene_count',
    'timepoint_numeric'),
  sort = FALSE)
tri21_displacement$tri21_minus_di21 <- tri21_displacement$maturation_score -
  tri21_displacement$d21_mean_score
tri21_displacement$timepoint_label <- base::factor(
  tri21_displacement$timepoint_label,
  levels = timepoint_levels)

ggplot2::ggplot(
  tri21_displacement,
  ggplot2::aes(
    effective_gene_count,
    tri21_minus_di21,
    color = timepoint_label,
    group = timepoint_label)) +
  ggplot2::geom_hline(yintercept = 0, linetype = 'dashed', linewidth = 0.3) +
  ggplot2::geom_line(linewidth = 0.65) +
  ggplot2::geom_point(size = 0.9) +
  ggplot2::facet_wrap(~interval, scales = 'free_y') +
  ggplot2::scale_x_continuous(breaks = sensitivity_x_breaks) +
  ggplot2::scale_color_manual(values = timepoint_colors) +
  ggplot2::labs(
    x = 'Effective number of input genes',
    y = 'Tri21 mean score minus matched Di21 mean score',
    title = 'Sensitivity of Tri21 displacement from matched Di21') +
  theme_ab01(base_size = 10) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

## Convergence Toward the Maximum-Gene Score

This plot compares each gene-count score with the score obtained when all
filtered VST genes are allowed into the interval-specific feature ranking. Lower
values indicate closer agreement with the maximum-gene score for the same
interval, genotype, and timepoint.

```{r sensitivity-score-convergence, fig.width=9.5, fig.height=5.0}
max_gene_count <- base::max(sensitivity_summary$effective_gene_count)
max_gene_reference <- sensitivity_summary[
  sensitivity_summary$effective_gene_count == max_gene_count,
  base::c(
    'interval',
    'genotype_display',
    'timepoint_numeric',
    'maturation_score'),
  drop = FALSE]
base::names(max_gene_reference)[
  base::names(max_gene_reference) == 'maturation_score'
] <- 'max_gene_maturation_score'

score_convergence <- base::merge(
  sensitivity_summary,
  max_gene_reference,
  by = base::c('interval', 'genotype_display', 'timepoint_numeric'),
  sort = FALSE)
score_convergence$absolute_difference_from_max_gene_score <- base::abs(
  score_convergence$maturation_score -
    score_convergence$max_gene_maturation_score)
score_convergence$label_group <- base::factor(
  base::paste(score_convergence$genotype_display, score_convergence$timepoint_label),
  levels = base::c(
    'Di21 D6',
    'Tri21 D6',
    'Di21 D10',
    'Tri21 D10',
    'Di21 D17',
    'Tri21 D17'))

ggplot2::ggplot(
  score_convergence,
  ggplot2::aes(
    effective_gene_count,
    absolute_difference_from_max_gene_score,
    color = genotype_display,
    linetype = timepoint_label,
    group = label_group)) +
  ggplot2::geom_hline(yintercept = 0, linetype = 'dashed', linewidth = 0.3) +
  ggplot2::geom_line(linewidth = 0.62) +
  ggplot2::geom_point(size = 0.9) +
  ggplot2::facet_wrap(~interval, scales = 'free_y') +
  ggplot2::scale_x_continuous(breaks = sensitivity_x_breaks) +
  ggplot2::scale_color_manual(values = genotype_colors) +
  ggplot2::labs(
    x = 'Effective number of input genes',
    y = 'Absolute difference from maximum-gene mean score',
    title = 'Convergence of interval scores as more genes enter the PCA') +
  theme_ab01(base_size = 10) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

# Summary

The summary table reports mean scores by interval, genotype, and timepoint. D21
means define the anchors by construction; the T21 values show where Tri21 samples
fall in the same D21-trained PCA coordinate system.

```{r final-summary}
primary_mean_summary <- stats::aggregate(
  maturation_score ~ interval + genotype_display + timepoint_label,
  data = primary_scores,
  FUN = base::mean)

primary_mean_summary <- primary_mean_summary[
  base::order(
    primary_mean_summary$interval,
    primary_mean_summary$timepoint_label,
    primary_mean_summary$genotype_display),
  ,
  drop = FALSE]

primary_mean_summary
```

# ZRN01 Mesenchyme PCA Checks

These final PCA checks use the copied ZRN01 objects stored under `data_DSHC1/`.
Respiratory mesenchyme and esophageal mesenchyme are analyzed separately. For
each lineage, genes are ranked by VST variance across that lineage's samples
only, the top 1000 genes are used for PCA, and the scree plot reports PC1-PC15.

```{r zrn01-load-data}
ZRN01_metadata <- base::readRDS('data_DSHC1/ZRN01/metadata/ZRN01_metadata.rds')
ZRN01_vst <- base::readRDS('data_DSHC1/ZRN01/counts/ZRN01_vst.rds')

base::stopifnot(!base::anyDuplicated(ZRN01_metadata$sampleid))
base::stopifnot(base::all(ZRN01_metadata$sampleid %in% base::colnames(ZRN01_vst)))

ZRN01_metadata <- ZRN01_metadata[
  base::match(base::colnames(ZRN01_vst), ZRN01_metadata$sampleid),
  ,
  drop = FALSE]
base::stopifnot(base::identical(
  base::as.character(ZRN01_metadata$sampleid),
  base::colnames(ZRN01_vst)))

ZRN01_metadata$timepoint_label <- base::factor(
  base::as.character(ZRN01_metadata$timepoint),
  levels = base::paste0('D', base::sort(base::unique(ZRN01_metadata$timepoint_numeric))))
ZRN01_metadata$genotype_display <- base::factor(
  base::as.character(ZRN01_metadata$genotype),
  levels = base::c('Di21', 'Tri21'))

base::data.frame(
  samples = base::nrow(ZRN01_metadata),
  lineages = base::paste(base::levels(ZRN01_metadata$lineage), collapse = ', '),
  genotypes = base::paste(base::levels(ZRN01_metadata$genotype), collapse = ', '),
  timepoints = base::paste(base::levels(ZRN01_metadata$timepoint_label), collapse = ', '),
  stringsAsFactors = FALSE)
```

```{r zrn01-pca-functions}
build_zrn01_lineage_pca <- function(vst_matrix, sample_metadata, lineage_name,
                                    requested_gene_count = pca_gene_count) {
  lineage_ids <- base::as.character(sample_metadata$sampleid[sample_metadata$lineage == lineage_name])
  gene_variance <- base::apply(vst_matrix[, lineage_ids, drop = FALSE], 1, stats::var)
  ranked_genes <- base::names(base::sort(gene_variance, decreasing = TRUE))
  effective_gene_count <- base::min(requested_gene_count, base::length(ranked_genes))
  gene_ids <- ranked_genes[base::seq_len(effective_gene_count)]

  pca_fit <- stats::prcomp(
    base::t(vst_matrix[gene_ids, lineage_ids, drop = FALSE]),
    center = TRUE,
    scale. = FALSE)
  coords <- base::as.data.frame(pca_fit$x)
  coords$sampleid <- base::rownames(coords)
  coords <- base::merge(
    sample_metadata,
    coords,
    by = 'sampleid',
    sort = FALSE)
  coords <- coords[base::match(lineage_ids, coords$sampleid), , drop = FALSE]

  base::list(
    pca_fit = pca_fit,
    coords = coords,
    gene_count = effective_gene_count,
    variance_explained = calculate_variance_explained(pca_fit))
}

plot_zrn01_pca <- function(pca_result, plot_title) {
  ggplot2::ggplot(
    pca_result$coords,
    ggplot2::aes(PC1, PC2, color = timepoint_label, shape = genotype_display)) +
    ggplot2::geom_point(size = 2.7, stroke = 0.8) +
    ggplot2::scale_color_manual(values = timepoint_colors) +
    ggplot2::scale_shape_manual(values = base::c(Di21 = 16, Tri21 = 17)) +
    ggplot2::labs(
      x = format_pc_axis_label('PC1', pca_result$variance_explained),
      y = format_pc_axis_label('PC2', pca_result$variance_explained),
      title = plot_title) +
    theme_ab01(base_size = 10)
}
```

## ZRN01 Respiratory Mesenchyme

```{r zrn01-rm-pca}
zrn01_rm_pca <- build_zrn01_lineage_pca(
  vst_matrix = ZRN01_vst,
  sample_metadata = ZRN01_metadata,
  lineage_name = 'RM',
  requested_gene_count = pca_gene_count)

base::data.frame(
  lineage = 'RM',
  effective_gene_count = zrn01_rm_pca$gene_count,
  samples = base::nrow(zrn01_rm_pca$coords),
  stringsAsFactors = FALSE)
```

```{r zrn01-rm-pca-plot, fig.width=5.6, fig.height=4.4}
plot_zrn01_pca(
  pca_result = zrn01_rm_pca,
  plot_title = 'ZRN01 respiratory mesenchyme PCA')
```

```{r zrn01-rm-scree-plot, fig.width=6.2, fig.height=4.2}
plot_scree(
  pca_result = zrn01_rm_pca,
  plot_title = 'ZRN01 respiratory mesenchyme scree plot')
```

## ZRN01 Esophageal Mesenchyme

```{r zrn01-em-pca}
zrn01_em_pca <- build_zrn01_lineage_pca(
  vst_matrix = ZRN01_vst,
  sample_metadata = ZRN01_metadata,
  lineage_name = 'EM',
  requested_gene_count = pca_gene_count)

base::data.frame(
  lineage = 'EM',
  effective_gene_count = zrn01_em_pca$gene_count,
  samples = base::nrow(zrn01_em_pca$coords),
  stringsAsFactors = FALSE)
```

```{r zrn01-em-pca-plot, fig.width=5.6, fig.height=4.4}
plot_zrn01_pca(
  pca_result = zrn01_em_pca,
  plot_title = 'ZRN01 esophageal mesenchyme PCA')
```

```{r zrn01-em-scree-plot, fig.width=6.2, fig.height=4.2}
plot_scree(
  pca_result = zrn01_em_pca,
  plot_title = 'ZRN01 esophageal mesenchyme scree plot')
```
