---
title: "AB01 Interval Maturation Scoring"
description: "A standalone test of adjacent-interval PCA maturation scoring in AB01."
output:
  html_document:
    toc: true
    toc_depth: 2
    code_folding: none
    df_print: paged
---

```{r setup}
# Created:
# 2026-06-03
#
# Inputs:
# - data/AB01/metadata/WC24_metadata_clean.rds: AB01 sample metadata
# - data/AB01/raw counts/WC24_filt_raw_counts.rds: AB01 raw count matrix
#
# Outputs:
# - docs/AB01_interval_maturation_score.html: rendered standalone HTML report
#
# Purpose:
# Test an adjacent-interval PCA maturation score for AB01. Each interval is
# trained on D21 samples only, using the top genes with the largest absolute
# D21 timepoint mean VST difference. T21 samples are projected into the fixed
# D21-trained PCA space and scored relative to the D21 interval centroid vector.
#
# Notes:
# VST is computed exactly once with DESeq2::vst(blind = FALSE) using the
# genotype_timepoint condition after CPM-based expression filtering.

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = TRUE,
  fig.align = 'center',
  dpi = 600
)

required_packages <- base::c(
  'DESeq2',
  'SummarizedExperiment',
  'ggplot2'
)
missing_packages <- required_packages[
  !base::vapply(
    required_packages,
    base::requireNamespace,
    base::logical(1),
    quietly = TRUE
  )
]
if (base::length(missing_packages) > 0L) {
  base::stop(
    'missing required packages: ',
    base::paste(missing_packages, collapse = ', ')
  )
}

plot_font_family <- 'Arial'

theme_ab01 <- function(base_size = 10) {
  ggplot2::theme_classic(
    base_size = base_size,
    base_family = plot_font_family
  ) +
    ggplot2::theme(
      text = ggplot2::element_text(color = 'black', family = plot_font_family),
      axis.text = ggplot2::element_text(color = 'black', family = plot_font_family),
      axis.title = ggplot2::element_text(color = 'black', family = plot_font_family),
      axis.line = ggplot2::element_line(color = 'black', linewidth = 0.3),
      axis.ticks = ggplot2::element_line(color = 'black', linewidth = 0.3),
      panel.grid.major.y = ggplot2::element_line(color = 'grey92', linewidth = 0.2),
      panel.grid.major.x = ggplot2::element_blank(),
      panel.grid.minor = ggplot2::element_blank(),
      strip.background = ggplot2::element_blank(),
      strip.text = ggplot2::element_text(
        color = 'black',
        face = 'bold',
        family = plot_font_family
      ),
      legend.title = ggplot2::element_blank(),
      legend.text = ggplot2::element_text(color = 'black', family = plot_font_family)
    )
}
```

# Overview

This report tests an adjacent-interval maturation score in AB01. The workflow
uses D21 as the reference trajectory and scores T21 samples relative to two
separate D21-trained intervals:

- D6 to D10, anchored so D21 D6 = 0 and D21 D10 = 1.
- D10 to D17, anchored so D21 D10 = 1 and D21 D17 = 2.

Each interval gets its own gene set and PCA model. The intervals are therefore
independent scoring spaces, even though D10 appears in both intervals.

# Load AB01 Data

```{r load-data}
metadata <- base::readRDS('data/AB01/metadata/WC24_metadata_clean.rds')
counts <- base::readRDS('data/AB01/raw counts/WC24_filt_raw_counts.rds')

base::stopifnot(!base::anyDuplicated(metadata$sample_id))
base::stopifnot(base::all(metadata$sample_id %in% base::colnames(counts)))

counts <- counts[, metadata$sample_id, drop = FALSE]
base::stopifnot(base::identical(
  base::as.character(metadata$sample_id),
  base::colnames(counts)
))

metadata$timepoint_numeric <- base::as.numeric(base::as.character(metadata$timepoint))
timepoint_levels <- base::paste0(
  'D',
  base::sort(base::unique(metadata$timepoint_numeric))
)
metadata$timepoint_label <- base::factor(
  base::paste0('D', metadata$timepoint_numeric),
  levels = timepoint_levels
)
metadata$genotype <- base::factor(metadata$genotype, levels = base::c('D21', 'T21'))
metadata$genotype_timepoint <- base::factor(
  metadata$genotype_timepoint,
  levels = base::unique(metadata$genotype_timepoint[
    base::order(metadata$timepoint_numeric, metadata$genotype)
  ])
)
base::rownames(metadata) <- metadata$sample_id

input_summary <- base::data.frame(
  samples = base::nrow(metadata),
  genotypes = base::paste(base::levels(metadata$genotype), collapse = ', '),
  timepoints = base::paste(
    base::sort(base::unique(metadata$timepoint_numeric)),
    collapse = ', '
  ),
  stringsAsFactors = FALSE
)

input_summary
```

# Preprocess Counts and Compute VST Once

This preprocessing section applies one expression filter before VST:
retain genes with CPM >= 1 in at least 3 samples. The filtered count matrix is
then used to generate one VST matrix, and that same VST matrix is reused for
every interval and every sensitivity-test gene count.

```{r compute-vst}
library_sizes <- base::colSums(counts)
cpm_matrix <- base::sweep(counts, 2, library_sizes, FUN = '/') * 1e6
keep_genes <- base::rowSums(cpm_matrix >= 1) >= 3
filtered_counts <- counts[keep_genes, , drop = FALSE]

preprocessing_summary <- base::data.frame(
  retained_genes_after_cpm_filter = base::nrow(filtered_counts),
  cpm_cutoff = 1,
  minimum_samples = 3,
  stringsAsFactors = FALSE
)

preprocessing_summary

dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = filtered_counts,
  colData = metadata,
  design = ~genotype_timepoint
)

vst_matrix <- SummarizedExperiment::assay(
  DESeq2::vst(dds, blind = FALSE)
)

base::stopifnot(base::identical(base::colnames(vst_matrix), metadata$sample_id))
```

# Define Interval Scoring Functions

```{r define-functions}
interval_specs <- base::data.frame(
  interval = base::c('D6-D10', 'D10-D17'),
  start_timepoint = base::c(6, 10),
  end_timepoint = base::c(10, 17),
  score_start = base::c(0, 1),
  score_end = base::c(1, 2),
  stringsAsFactors = FALSE
)
interval_specs$interval <- base::factor(
  interval_specs$interval,
  levels = base::c('D6-D10', 'D10-D17')
)

sensitivity_gene_counts <- base::unique(base::c(
  base::seq(100L, base::nrow(vst_matrix), by = 100L),
  base::nrow(vst_matrix)
))
sensitivity_x_breaks <- base::seq(500L, base::max(sensitivity_gene_counts), by = 500L)

get_interval_sample_ids <- function(sample_metadata, start_timepoint, end_timepoint, genotype = NULL) {
  keep <- sample_metadata$timepoint_numeric %in% base::c(start_timepoint, end_timepoint)
  if (!base::is.null(genotype)) {
    keep <- keep & sample_metadata$genotype == genotype
  }
  base::as.character(sample_metadata$sample_id[keep])
}

select_top_interval_genes <- function(vst_matrix, sample_metadata,
                                      start_timepoint, end_timepoint,
                                      requested_gene_count) {
  start_ids <- get_interval_sample_ids(
    sample_metadata,
    start_timepoint = start_timepoint,
    end_timepoint = start_timepoint,
    genotype = 'D21'
  )
  end_ids <- get_interval_sample_ids(
    sample_metadata,
    start_timepoint = end_timepoint,
    end_timepoint = end_timepoint,
    genotype = 'D21'
  )

  # Rank genes by the absolute difference between D21 timepoint means.
  # This intentionally targets between-timepoint change instead of
  # replicate-to-replicate spread within either timepoint.
  start_mean <- base::rowMeans(vst_matrix[, start_ids, drop = FALSE])
  end_mean <- base::rowMeans(vst_matrix[, end_ids, drop = FALSE])
  abs_mean_delta <- base::abs(end_mean - start_mean)
  ranked_genes <- base::names(base::sort(abs_mean_delta, decreasing = TRUE))
  effective_gene_count <- base::min(requested_gene_count, base::length(ranked_genes))

  base::data.frame(
    gene_id = ranked_genes[base::seq_len(effective_gene_count)],
    abs_mean_delta = abs_mean_delta[ranked_genes[base::seq_len(effective_gene_count)]],
    requested_gene_count = requested_gene_count,
    effective_gene_count = effective_gene_count,
    stringsAsFactors = FALSE
  )
}

project_samples_into_pca <- function(vst_subset, pca_fit) {
  if (!base::identical(base::rownames(vst_subset), base::rownames(pca_fit$rotation))) {
    base::stop('vst_subset rows must match pca_fit rotation rows exactly.')
  }

  projected <- base::scale(
    base::t(vst_subset),
    center = pca_fit$center,
    scale = FALSE
  )
  coord <- base::as.data.frame(projected %*% pca_fit$rotation)
  coord$sample_id <- base::rownames(coord)
  coord
}

score_interval <- function(vst_matrix, sample_metadata, interval_row,
                           requested_gene_count = 1000L,
                           n_scoring_pcs = 3L) {
  start_timepoint <- interval_row$start_timepoint[[1]]
  end_timepoint <- interval_row$end_timepoint[[1]]
  score_start <- interval_row$score_start[[1]]
  score_end <- interval_row$score_end[[1]]

  gene_table <- select_top_interval_genes(
    vst_matrix = vst_matrix,
    sample_metadata = sample_metadata,
    start_timepoint = start_timepoint,
    end_timepoint = end_timepoint,
    requested_gene_count = requested_gene_count
  )
  gene_ids <- gene_table$gene_id

  training_ids <- get_interval_sample_ids(
    sample_metadata,
    start_timepoint = start_timepoint,
    end_timepoint = end_timepoint,
    genotype = 'D21'
  )
  projection_ids <- get_interval_sample_ids(
    sample_metadata,
    start_timepoint = start_timepoint,
    end_timepoint = end_timepoint
  )

  pca_fit <- stats::prcomp(
    base::t(vst_matrix[gene_ids, training_ids, drop = FALSE]),
    center = TRUE,
    scale. = FALSE
  )
  pc_cols <- base::paste0(
    'PC',
    base::seq_len(base::min(n_scoring_pcs, base::ncol(pca_fit$x)))
  )

  coords <- project_samples_into_pca(
    vst_subset = vst_matrix[gene_ids, projection_ids, drop = FALSE],
    pca_fit = pca_fit
  )
  coords <- coords[, base::c('sample_id', pc_cols), drop = FALSE]
  score_df <- base::merge(
    sample_metadata,
    coords,
    by = 'sample_id',
    sort = FALSE
  )
  score_df <- score_df[base::match(projection_ids, score_df$sample_id), , drop = FALSE]

  start_centroid <- base::colMeans(score_df[
    score_df$genotype == 'D21' & score_df$timepoint_numeric == start_timepoint,
    pc_cols,
    drop = FALSE
  ])
  end_centroid <- base::colMeans(score_df[
    score_df$genotype == 'D21' & score_df$timepoint_numeric == end_timepoint,
    pc_cols,
    drop = FALSE
  ])
  vector_raw <- end_centroid - start_centroid
  vector_length_sq <- base::sum(vector_raw^2)
  if (!base::is.finite(vector_length_sq) || vector_length_sq <= 0) {
    base::stop('interval centroid vector has zero length.')
  }

  coords_mat <- base::as.matrix(score_df[, pc_cols, drop = FALSE])
  centered_coords <- coords_mat - base::matrix(
    start_centroid,
    nrow = base::nrow(coords_mat),
    ncol = base::length(pc_cols),
    byrow = TRUE
  )
  unit_interval_score <- base::as.numeric(centered_coords %*% vector_raw) / vector_length_sq
  score_df$maturation_score <- score_start +
    unit_interval_score * (score_end - score_start)

  projected_coords <- base::matrix(
    start_centroid,
    nrow = base::nrow(coords_mat),
    ncol = base::length(pc_cols),
    byrow = TRUE
  ) + unit_interval_score %o% vector_raw
  for (pc_idx in base::seq_along(pc_cols)) {
    score_df[[base::paste0('vector_', pc_cols[[pc_idx]])]] <- projected_coords[, pc_idx]
  }

  score_df$interval <- interval_row$interval[[1]]
  score_df$interval <- base::factor(
    score_df$interval,
    levels = base::levels(interval_specs$interval)
  )
  score_df$interval_start_timepoint <- start_timepoint
  score_df$interval_end_timepoint <- end_timepoint
  score_df$score_start <- score_start
  score_df$score_end <- score_end
  score_df$requested_gene_count <- requested_gene_count
  score_df$effective_gene_count <- gene_table$effective_gene_count[[1]]

  base::list(
    score_table = score_df,
    gene_table = gene_table,
    pca_fit = pca_fit,
    pc_columns = pc_cols,
    start_centroid = start_centroid,
    end_centroid = end_centroid
  )
}
```

# Run the Primary 1000-Gene Interval Score

```{r run-primary-score}
primary_results <- base::lapply(
  base::seq_len(base::nrow(interval_specs)),
  function(i) {
    score_interval(
      vst_matrix = vst_matrix,
      sample_metadata = metadata,
      interval_row = interval_specs[i, , drop = FALSE],
      requested_gene_count = 1000L
    )
  }
)
base::names(primary_results) <- interval_specs$interval

primary_scores <- base::do.call(
  base::rbind,
  base::lapply(primary_results, function(x) x$score_table)
)
base::rownames(primary_scores) <- NULL

primary_scores[
  ,
  base::c(
    'interval',
    'sample_id',
    'genotype',
    'timepoint_label',
    'rep',
    'PC1',
    'PC2',
    'PC3',
    'maturation_score'
  ),
  drop = FALSE
]
```

# Identify Top Interval-Variable Genes

Genes are ranked separately for each interval using D21 samples only. For each
gene, the D21 mean VST value at the start timepoint is subtracted from the D21
mean VST value at the end timepoint. Genes with the largest absolute mean
difference define that interval's PCA input list.

```{r interval-variable-genes}
primary_gene_summary <- base::do.call(
  base::rbind,
  base::lapply(base::names(primary_results), function(interval_name) {
    gene_table <- primary_results[[interval_name]]$gene_table
    base::data.frame(
      interval = interval_name,
      requested_gene_count = gene_table$requested_gene_count[[1]],
      effective_gene_count = gene_table$effective_gene_count[[1]],
      largest_abs_mean_delta = base::max(gene_table$abs_mean_delta),
      stringsAsFactors = FALSE
    )
  })
)

primary_gene_summary
```

# Primary PCA and Score Views

```{r primary-pca-plot, fig.width=9, fig.height=4.4}
centroid_rows <- base::do.call(
  base::rbind,
  base::lapply(primary_results, function(interval_result) {
    score_df <- interval_result$score_table
    start_row <- base::data.frame(
      interval = score_df$interval[[1]],
      timepoint_numeric = score_df$interval_start_timepoint[[1]],
      PC1 = interval_result$start_centroid[['PC1']],
      PC2 = interval_result$start_centroid[['PC2']],
      stringsAsFactors = FALSE
    )
    end_row <- base::data.frame(
      interval = score_df$interval[[1]],
      timepoint_numeric = score_df$interval_end_timepoint[[1]],
      PC1 = interval_result$end_centroid[['PC1']],
      PC2 = interval_result$end_centroid[['PC2']],
      stringsAsFactors = FALSE
    )
    base::rbind(start_row, end_row)
  })
)

ggplot2::ggplot(
  primary_scores,
  ggplot2::aes(PC1, PC2, color = timepoint_label, shape = genotype)
) +
  ggplot2::geom_path(
    data = centroid_rows,
    ggplot2::aes(PC1, PC2, group = interval),
    inherit.aes = FALSE,
    color = 'black',
    linewidth = 0.55,
    arrow = grid::arrow(length = grid::unit(0.15, 'cm'), type = 'closed')
  ) +
  ggplot2::geom_point(size = 2.5, stroke = 0.8) +
  ggplot2::facet_wrap(~interval, scales = 'free') +
  ggplot2::scale_color_manual(values = base::c(D6 = '#059669', D10 = '#7C3AED', D17 = '#EA580C')) +
  ggplot2::scale_shape_manual(values = base::c(D21 = 16, T21 = 17)) +
  ggplot2::labs(
    x = 'PC1',
    y = 'PC2',
    title = 'Independent D21-trained interval PCA spaces'
  ) +
  theme_ab01(base_size = 10)
```

```{r primary-score-plot, fig.width=8.4, fig.height=4.4}
score_plot_df <- primary_scores[
  primary_scores$timepoint_numeric == primary_scores$interval_start_timepoint,
  ,
  drop = FALSE
]

score_mean <- stats::aggregate(
  maturation_score ~ interval + genotype + timepoint_label,
  data = score_plot_df,
  FUN = base::mean
)
score_mean$interval <- base::factor(
  score_mean$interval,
  levels = base::levels(interval_specs$interval)
)

ggplot2::ggplot(
  score_mean,
  ggplot2::aes(genotype, maturation_score, fill = genotype)
) +
  ggplot2::geom_hline(
    yintercept = base::c(0, 1, 2),
    linetype = 'dashed',
    color = 'black',
    linewidth = 0.25
  ) +
  ggplot2::geom_col(
    width = 0.62,
    color = 'black',
    linewidth = 0.25,
    alpha = 0.9
  ) +
  ggplot2::geom_point(
    data = score_plot_df,
    ggplot2::aes(genotype, maturation_score, fill = genotype),
    inherit.aes = FALSE,
    size = 1.9,
    shape = 21,
    color = 'black',
    stroke = 0.25,
    position = ggplot2::position_jitter(width = 0.06, height = 0, seed = 42)
  ) +
  ggplot2::facet_grid(. ~ interval) +
  ggplot2::scale_fill_manual(values = base::c(D21 = '#2563EB', T21 = '#DC2626')) +
  ggplot2::labs(
    x = NULL,
    y = 'Interval maturation score',
    title = 'Start-timepoint scores from each D21-trained interval'
  ) +
  theme_ab01(base_size = 10)
```

# Sensitivity Test

The sensitivity test repeats the same interval workflow every 100 genes from
100 genes to the maximum available AB01 gene count. X-axis tick labels are shown
every 500 genes.

```{r run-sensitivity}
sensitivity_results <- base::list()
result_idx <- 1L

for (interval_idx in base::seq_len(base::nrow(interval_specs))) {
  for (gene_count in sensitivity_gene_counts) {
    sensitivity_results[[result_idx]] <- score_interval(
      vst_matrix = vst_matrix,
      sample_metadata = metadata,
      interval_row = interval_specs[interval_idx, , drop = FALSE],
      requested_gene_count = gene_count
    )
    result_idx <- result_idx + 1L
  }
}

sensitivity_scores <- base::do.call(
  base::rbind,
  base::lapply(sensitivity_results, function(x) x$score_table)
)
base::rownames(sensitivity_scores) <- NULL

sensitivity_summary <- stats::aggregate(
  maturation_score ~ interval + requested_gene_count + effective_gene_count +
    genotype + timepoint_numeric + timepoint_label,
  data = sensitivity_scores,
  FUN = base::mean
)

sensitivity_summary <- sensitivity_summary[
  base::order(
    sensitivity_summary$interval,
    sensitivity_summary$requested_gene_count,
    sensitivity_summary$genotype,
    sensitivity_summary$timepoint_numeric
  ),
  ,
  drop = FALSE
]

utils::head(sensitivity_summary, 12)
```

## Sensitivity Plot 1: Mean Score Across Gene Counts

```{r sensitivity-mean-score, fig.width=9.5, fig.height=5.2}
ggplot2::ggplot(
  sensitivity_summary,
  ggplot2::aes(
    effective_gene_count,
    maturation_score,
    color = genotype,
    linetype = timepoint_label,
    group = base::interaction(genotype, timepoint_label)
  )
) +
  ggplot2::geom_hline(
    yintercept = base::c(0, 1, 2),
    linetype = 'dashed',
    color = 'grey45',
    linewidth = 0.25
  ) +
  ggplot2::geom_line(linewidth = 0.62) +
  ggplot2::geom_point(size = 0.9) +
  ggplot2::facet_wrap(~interval, scales = 'free_y') +
  ggplot2::scale_x_continuous(
    breaks = sensitivity_x_breaks
  ) +
  ggplot2::scale_color_manual(values = base::c(D21 = '#2563EB', T21 = '#DC2626')) +
  ggplot2::labs(
    x = 'Effective number of input genes',
    y = 'Mean interval maturation score',
    title = 'Sensitivity of genotype-timepoint mean scores to gene-list size'
  ) +
  theme_ab01(base_size = 10) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

## Sensitivity Plot 2: T21 Offset From Matched D21 Anchor

```{r sensitivity-t21-offset, fig.width=9, fig.height=4.8}
d21_reference <- sensitivity_summary[
  sensitivity_summary$genotype == 'D21',
  base::c(
    'interval',
    'requested_gene_count',
    'effective_gene_count',
    'timepoint_numeric',
    'maturation_score'
  ),
  drop = FALSE
]
base::names(d21_reference)[base::names(d21_reference) == 'maturation_score'] <- 'd21_mean_score'

t21_offset <- base::merge(
  sensitivity_summary[sensitivity_summary$genotype == 'T21', , drop = FALSE],
  d21_reference,
  by = base::c(
    'interval',
    'requested_gene_count',
    'effective_gene_count',
    'timepoint_numeric'
  ),
  sort = FALSE
)
t21_offset$t21_minus_d21 <- t21_offset$maturation_score - t21_offset$d21_mean_score
t21_offset$timepoint_label <- base::factor(
  t21_offset$timepoint_label,
  levels = timepoint_levels
)

ggplot2::ggplot(
  t21_offset,
  ggplot2::aes(
    effective_gene_count,
    t21_minus_d21,
    color = timepoint_label,
    group = timepoint_label
  )
) +
  ggplot2::geom_hline(yintercept = 0, linetype = 'dashed', linewidth = 0.3) +
  ggplot2::geom_line(linewidth = 0.65) +
  ggplot2::geom_point(size = 0.9) +
  ggplot2::facet_wrap(~interval, scales = 'free_y') +
  ggplot2::scale_x_continuous(
    breaks = sensitivity_x_breaks
  ) +
  ggplot2::scale_color_manual(values = base::c(D6 = '#059669', D10 = '#7C3AED', D17 = '#EA580C')) +
  ggplot2::labs(
    x = 'Effective number of input genes',
    y = 'T21 mean score minus matched D21 mean score',
    title = 'Sensitivity of T21 displacement from the D21 interval anchors'
  ) +
  theme_ab01(base_size = 10) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

## Sensitivity Plot 3: Convergence Toward the Maximum-Gene Score

```{r sensitivity-score-convergence, fig.width=9.5, fig.height=5.0}
max_gene_count <- base::max(sensitivity_summary$effective_gene_count)
max_gene_reference <- sensitivity_summary[
  sensitivity_summary$effective_gene_count == max_gene_count,
  base::c(
    'interval',
    'genotype',
    'timepoint_numeric',
    'maturation_score'
  ),
  drop = FALSE
]
base::names(max_gene_reference)[
  base::names(max_gene_reference) == 'maturation_score'
] <- 'max_gene_maturation_score'

score_convergence <- base::merge(
  sensitivity_summary,
  max_gene_reference,
  by = base::c('interval', 'genotype', 'timepoint_numeric'),
  sort = FALSE
)
score_convergence$absolute_difference_from_max_gene_score <- base::abs(
  score_convergence$maturation_score -
    score_convergence$max_gene_maturation_score
)
score_convergence$label_group <- base::factor(
  base::paste(score_convergence$genotype, score_convergence$timepoint_label),
  levels = base::c(
    'D21 D6',
    'T21 D6',
    'D21 D10',
    'T21 D10',
    'D21 D17',
    'T21 D17'
  )
)

ggplot2::ggplot(
  score_convergence,
  ggplot2::aes(
    effective_gene_count,
    absolute_difference_from_max_gene_score,
    color = genotype,
    linetype = timepoint_label,
    group = label_group
  )
) +
  ggplot2::geom_hline(yintercept = 0, linetype = 'dashed', linewidth = 0.3) +
  ggplot2::geom_line(linewidth = 0.62) +
  ggplot2::geom_point(size = 0.9) +
  ggplot2::facet_wrap(~interval, scales = 'free_y') +
  ggplot2::scale_x_continuous(
    breaks = sensitivity_x_breaks
  ) +
  ggplot2::scale_color_manual(values = base::c(D21 = '#2563EB', T21 = '#DC2626')) +
  ggplot2::labs(
    x = 'Effective number of input genes',
    y = 'Absolute difference from maximum-gene mean score',
    title = 'Convergence of interval scores as more genes enter the PCA'
  ) +
  theme_ab01(base_size = 10) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

# Summary

```{r final-summary}
primary_mean_summary <- stats::aggregate(
  maturation_score ~ interval + genotype + timepoint_label,
  data = primary_scores,
  FUN = base::mean
)

primary_mean_summary <- primary_mean_summary[
  base::order(
    primary_mean_summary$interval,
    primary_mean_summary$genotype,
    primary_mean_summary$timepoint_label
  ),
  ,
  drop = FALSE
]

primary_mean_summary
```
